refnx - Neutron and X-ray reflectometry analysis in Python

### 2018-07-19 paper/manuscript.tex
PAPER: manuscript changes

\newcommand{\refnx}{\emph{refnx}}
\newcommand{\Objective}{\textbf{Objective}}
\newcommand{\Parameter}{\textbf{Parameter}}
\newcommand{\Structure}{\textbf{Structure}}
\newcommand{\Slab}{\textbf{Slab}}
\newcommand{\Component}{\textbf{Component}}
\newcommand{\LipidLeaflet}{\textbf{LipidLeaflet}}
\newcommand{\DataD}{\textbf{Data1D}}
\newcommand{\ReflectModel}{\textbf{ReflectModel}}
\newcommand{\Spline}{\textbf{Spline}}
\newcommand{\conda}{\emph{conda}}
\newcommand{\MixedReflectModel}{\emph{MixedReflectModel}}
\newcommand{\pip}{\emph{pip}}
\newcommand{\NumPy}{\emph{NumPy}}
\newcommand{\SciPy}{\emph{SciPy}}
\newcommand{\Cython}{\emph{Cython}}
\newcommand{\Jupyter}{\emph{Jupyter}}
\newcommand{\ipywidgets}{\emph{ipywidgets}}

@@

\refnx\ has a wide range of sophisticated modelling capabilities including: parameter constraints, description of prior probabilities on parameters using arbitrary probability distribution functions, modular construction of models using a variety of components that are user extensible, co-refinement of datasets, handling of patchy layers, a graphical user interface, resolution handling, Markov Chain Monte Carlo sampling of the posterior probability distributions, model selection, etc.

@@

\begin{gather} 
\label{eqn:1}\ p(\theta | D, I) \propto p(\theta | I)\times p(D | \theta, I)\\
p(D | \theta, I) = -\frac{1}{2} \sum [(\frac{y_n - model} {\sigma_n})^2 + log(\sigma_n^2)]\label{eqn:2}
\end{gather}

The posterior distribution, $p(\theta | I, D)$, is the probability of the parameters given the data, $D$, and prior knowledge about the system, $I$. 
The prior probability distribution, $p(\theta | I)$, is the state of knowledge that already exists about the parameters of interest, such as that obtained from other experiments.
The likelihood (Equation~\ref{eqn:2}), $p(D | \theta, I)$, is the probability of the observed data, $y_n$ (with uncertainties $\sigma_n$), given the model parameters and other prior information.  The posterior probability is sampled by encoding the likelihood and prior distributions and using Markov Chain Monte Carlo (MCMC) techniques \cite{emcee}.

\section*{Method}
\refnx\ is written in Python with an object-oriented design, Figure~\ref{fig:components}. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.
The building block of the analysis is the \Parameter\ object which represents a model value, whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution - the simplest bound being a uniform distribution which would set lower and upper limits for the parameter. This probability distribution is the prior probability for the parameter. Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter\ can be applied using constraints.

@@

The \Objective\ class uses the \ReflectModel\ and a dataset, \DataD, to calculate $\chi^2$, log-likelihood, log-prior, residuals, and the generative model.
The \DataD object has \emph{x, y, y\_err, x\_err} attributes to represent $Q, R, dR, dQ$. The \DataD object can be constructed from a tuple of arrays representing the data, or by reading a three or four column plain-text datafile. A three column dataset represents Q ($\AA^{-1}$), R, dR (1 standard deviation). A four column dataset represents Q, R, dR, dQ ($\AA^{-1}$). Subclassing \DataD would allow other formats to be read. One example of this could be an wavelength dispersive file using $\Omega, \lambda$ data instead of Q; such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers. In such a case \ReflectModel\ could be subclassed to make full use of this information.
The \Objective\ can be given a \textbf{Transform} object to permit fitting as log-R vs Q, $RQ^4$ vs Q, or lin-R vs Q. Several \Objective\ can be combined to form a \textbf{GlobalObjective} for co-refinement. The object-oriented nature allows \Parameter\, \Component\ to be reused throughout the setup, thereby facilitating co-refinement.

The \Objective\ statistics are used directly by the \textbf{CurveFitter} class to perform least-square fitting or Bayesian MCMC of the system. \refnx\ uses the \emph{scipy} package for least-squares analysis (DifferentialEvolution, Levenberg Marquardt, LBFGSB - Limited Broyden–Fletcher–Goldfarb–Shanno with bounds) and the  \emph{emcee} package \cite{emcee} for affine invariant MCMC ensemble sampling of the posterior distribution of the parameters. The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective\  overriding the log-likelihood method.
Parallelisation of the sampling is done automatically, and can use MPI on a cluster for greater parallelisation. Visualisation of the samples produced by MCMC sampling is done using the \emph{corner} package \cite{corner} for scatter plot matrices. On the conclusion of a fit/sampling run each \Parameter\ is updated and given a standard error. For the sampling these represent the median and half the [15.87, 84.13] percentile range respectively; the latter approximates the standard deviation for a normally distributed statistic.

@@

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting graphical user interface (GUI) can be created in a \Jupyter\ notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the 'To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}. For a comprehensive demonstration of multiple contrast co-refinement see \textbf{lipid.ipynb}, \textbf{lipid.pdf} in the supporting information.

@@

\subsection*{Example}
Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel\. It is straightforward to develop/modify new components for different structural functionality, a consequence of the object-oriented approach. Here the \LipidLeaflet\ is used to co-refine three contrasts (\ce{D_2O}, \ce{Si} contrast match [hdmix, SLD=$2.07\times10^{-6}\AA^{-2}$], and \ce{H_2O}) of a 1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC) bilayer at the solid-liquid interface, Figure \ref{fig:global_fit} \footnote{The validity of \LipidLeaflet\ does depend on the area per molecule being equal for the headgroup and tailgroup regions, as pointed out by Gerelli \cite{Gerelli2016}, which can be violated if there are guest molecules that insert in the membrane.}. The corner plot (Figure \ref{fig:corner}) produced from the MCMC analysis shows the covariance between parameters, with an area per molecule of 57$\AA\pm1$. Figures~\ref{fig:global_fit} and \ref{fig:d2o_sld_spread} show the probability distribution of the generative model around the data and in the SLD profile. These are obtained by random samples from the MCMC chain. The spread in SLD profiles is used to determine what range of structures is consistent with the data. Multi-modalities in these SLD profiles can be due to statistical uncertainties, the Q ranges measured, and the loss of phase information in NR \cite{Majkrzak1999, Heinrich2009}.
%TODO use exact area per molecule, and use a reference for the Xray analysis

\begin{figure}%
\centering
\subfigure{%
\label{fig:global_fit}%
\includegraphics[width=5in]{./supporting_information/global_fit.png}}\hspace{1em}%
\subfigure{%
\label{fig:d2o_sld_spread}%
\includegraphics[width=5in]{./supporting_information/d2o_sld_spread.png}}\\%
\caption{a) Reflectivity from a DMPC bilayer measured at three contrasts, with median in red, and 500 samples from the posterior distribution. Dataset for \ce{Si} match offset by 0.1, \ce{H_2O} contrast offset by 0.001. b) SLD profile of the \ce{D_2O} model showing 500 samples from the posterior distribution, as well as the median. \Jupyter\ notebook used for the analysis is attached in the supporting information.}
\end{figure}

@@

\section*{Conclusions}\label{conclusions}
\refnx\ is a powerful tool for least-squares or Bayesian analysis of neutron and X-ray reflectometry data that is ideally usable for reproducible research with \Jupyter\ notebooks. Its features include: MCMC sampling of posterior distribution for parameters, structural models constructed from modular components with physically relevant parameterisation, interparameter constraints, mixed area models, co-refinement of multiple datasets, probability distributions for parameter bounds used directly for log-prior terms, and a \Jupyter\/\ipywidgets\ GUI

\section*{Acknowledgements}
We acknowledge Anton Le Brun (ANSTO) for the provision of the lipid bilayer datasets in the example, James Hester (ANSTO) for comments made on the draft manuscript, and Isaac Gresham (UNSW) for important feedback on refnx development.

>>>>

\newcommand{\refnx}{\emph{refnx}}
\newcommand{\Objective}{\textbf{Objective}}
\newcommand{\GlobalObjective}{\textbf{GlobalObjective}}
\newcommand{\Parameter}{\textbf{Parameter}}
\newcommand{\Structure}{\textbf{Structure}}
\newcommand{\Slab}{\textbf{Slab}}
\newcommand{\Component}{\textbf{Component}}
\newcommand{\LipidLeaflet}{\textbf{LipidLeaflet}}
\newcommand{\Transform}{\textbf{Transform}}
\newcommand{\DataD}{\textbf{Data1D}}
\newcommand{\ReflectModel}{\textbf{ReflectModel}}
\newcommand{\Spline}{\textbf{Spline}}
\newcommand{\conda}{\emph{conda}}
\newcommand{\corner}{\emph{corner}}
\newcommand{\MixedReflectModel}{\emph{MixedReflectModel}}
\newcommand{\pip}{\emph{pip}}
\newcommand{\emcee}{\emph{emcee}}
\newcommand{\NumPy}{\emph{NumPy}}
\newcommand{\SciPy}{\emph{SciPy}}
\newcommand{\Cython}{\emph{Cython}}
\newcommand{\Jupyter}{\emph{Jupyter}}
\newcommand{\ipywidgets}{\emph{ipywidgets}}

@@

\refnx\ has a wide range of sophisticated modelling capabilities including: parameter constraints, description of prior probabilities on parameters using arbitrary probability distribution functions, modular construction of models using a variety of components that are user extensible, co-refinement of datasets, handling of patchy layers, a graphical user interface, resolution handling, Markov Chain Monte Carlo sampling of the posterior probability distributions for the parameters, model selection, etc.

@@

\begin{gather} 
\label{eqn:1}\ p(\theta | D, I) \propto p(\theta | I)\times p(D | \theta, I)\\
p(D | \theta, I) = -\frac{1}{2} \sum \left[\left(\frac{y_n - model} {\sigma_n}\right)^2 + log(\sigma_n^2)\right]\label{eqn:2}
\end{gather}

The prior distribution, $p(\theta | I)$, is the probability of a parameter, $\theta$, given preexisting knowledge of the system, $I$. For example, a prior could be a simple Uniform distribution that specifies a lower and upper bound, or a normal distribution that represents an experimentally derived value.
The likelihood (Equation~\ref{eqn:2}), $p(D | \theta, I)$, is the probability of the observed data, $D$, given the model parameters and other prior information. It is calculated from the measured data, $y_n$ (with uncertainties $\sigma_n$), and the generative model, $model$.

The posterior distribution, $p(\theta | I, D)$, is the probability of the parameters given the data and prior knowledge about the system.  The posterior probability is sampled by encoding the likelihood and prior distributions and using Markov Chain Monte Carlo (MCMC) techniques \cite{emcee}. At the end of an MCMC run the parameter set possesses a number of samples (called a 'chain'), whose statistics reveal the distribution and covariance of the parameters, the spread of the model around the data, and in a reflectometry context the range of structures that are consistent with the data.

\section*{Method}
\refnx\ is written in Python with an object-oriented design, Figure~\ref{fig:components}. As with \emph{Motofit} \cite{Nelson2006} it calculates reflectivity using the Abeles method \cite{Heavens1955} for specular reflection from a stratified medium.
The building block of the analysis is the \Parameter\ object which represents a model value, whether it is allowed to vary in a fit, and a bounds attribute. The bounds are a probability distribution representing the prior. Any of the \emph{scipy.stats} \cite{Jones2001-2017} continuous distributions, or distributions created by the user, can be used for this purpose. Algebraic relationships between \Parameter\ can be applied using constraints.

@@

The \Objective\ class uses the \ReflectModel\ and a dataset, \DataD, to calculate $\chi^2$, log-likelihood, log-prior, residuals, and the generative model.
The \DataD\ object has \emph{x, y, y\_err, x\_err} attributes to represent $Q, R, dR, dQ$. The \DataD\ object can be constructed from a tuple of arrays representing the data, or by reading a three or four column plain-text datafile. A three column dataset represents Q ($\AA^{-1}$), R, dR (1 standard deviation). A four column dataset represents Q, R, dR, dQ ($\AA^{-1}$). Subclassing \DataD\ would allow other formats to be read. One example of this could be an wavelength dispersive file using $\Omega, \lambda$ data instead of Q; such as that used in energy scanned X-ray reflectometry, or sometimes produced by wavelength dispersive neutron reflectometers. In such a case \ReflectModel\ could be subclassed to make full use of this energy dispersive information.
An \Objective\ can be given a \Transform\ object to permit fitting as log-R vs Q, $RQ^4$ vs Q, or lin-R vs Q. Several \Objective\ can be combined to form a \GlobalObjective\ for co-refinement. The object-oriented nature allows \Parameter\, \Component\ to be reused in in different \Objective, thereby facilitating co-refinement.

The \Objective\ statistics are used directly by the \textbf{CurveFitter} class to perform least-square fitting or Bayesian MCMC of the system. \refnx\ uses the \SciPy\ package for least-squares analysis (DifferentialEvolution, Levenberg Marquardt, LBFGSB - Limited Broyden–Fletcher–Goldfarb–Shanno with bounds) and the  \emcee\ package \cite{emcee} for affine invariant MCMC ensemble sampling of the posterior distribution of the parameters. The likelihoods used in MCMC and least-squares assume that the measurement errors are normally distributed, Equation~\ref{eqn:2}. However, other types of measurement errors (e.g. Poissonian) could be implemented by a subclass of \Objective\  overriding the log-likelihood method.
Parallelisation of the sampling is done automatically, and can use MPI on a cluster for greater parallelisation. Visualisation of the samples produced by MCMC sampling is done using the \corner\ package \cite{corner} for scatter plot matrices. On the conclusion of a fit/sampling run each \Parameter\ value is updated and given a standard error. For the sampling these represent the median and half the [15.87, 84.13] percentile range respectively; the latter approximating the standard deviation for a normally distributed statistic.

@@

To make it easy to create quick models, and to ease the learning curve, an ipywidgets \cite{ipywidgets} based fitting graphical user interface (GUI) can be created in a \Jupyter\ notebook, Figure~\ref{fig:gui}. This GUI builds slab based models for analysing a single dataset, but can be used as a stepping point for building more advanced models via the 'To code' code generation button. A fully functional web-based reflectometry analysis notebook is currently available \cite{Nelson2018}. For a comprehensive demonstration of multiple contrast co-refinement see the annotated notebook in the supporting information.

@@

\subsection*{Example}
Neutron reflectometry is an ideal technique for the study of biologically relevant lipid membrane mimics and their interactions with proteins, etc. Multiple contrast variation measurements are necessary to reduce modelling ambiguity (due to loss of phase information in the scattering experiment) and improve the ability to determine the structure of various components in the system. The gold standard approach for analysis of these datasets is co-refinement with a common model, and to parameterise the model in terms of chemically relevant parameters, such as area per molecule. Sometimes a patchy coverage (distinct to low area per molecule) necessitates the use of an (incoherent) sum of reflectivities from different areas. \refnx\ has functionality for all these requirements, such as the \LipidLeaflet\ component for describing the head and tail groups of a lipid leaflet, and \MixedReflectModel\. It is straightforward to develop/modify new components for different structural functionality, a consequence of the object-oriented approach. Here the \LipidLeaflet\ is used to co-refine three contrasts (\ce{D_2O}, \ce{Si} contrast match [hdmix, SLD=$2.07\times10^{-6}\AA^{-2}$], and \ce{H_2O}) of a 1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC) bilayer at the solid-liquid interface, Figure \ref{fig:global_fit} \footnote{The validity of \LipidLeaflet\ does depend on the area per molecule being equal for the headgroup and tailgroup regions, as pointed out by Gerelli \cite{Gerelli2016}, which can be violated if there are guest molecules that insert in the membrane.}. The \Jupyter\ notebook used for the analysis, \emph{lipid.ipynb}, is available in the supplementary information. The corner plot (Figure \ref{fig:corner}) produced from the MCMC analysis shows the covariance between parameters, with an area per molecule of 57$\AA\pm1$. Figures~\ref{fig:global_fit} and \ref{fig:d2o_sld_spread} show the probability distribution of the generative model around the data and in the SLD profile. These are obtained by random samples from the MCMC chain. The spread in SLD profiles is used to determine what range of structures is consistent with the data. Multi-modalities in these SLD profiles can be due to statistical uncertainties, the Q ranges measured, and the loss of phase information in NR \cite{Majkrzak1999, Heinrich2009}.
%TODO use exact area per molecule, and use a reference for the Xray analysis

\begin{figure}%
\centering
\subfigure{%
\label{fig:global_fit}%
\includegraphics[width=5in]{./supporting_information/global_fit.png}}\hspace{1em}%
\subfigure{%
\label{fig:d2o_sld_spread}%
\includegraphics[width=5in]{./supporting_information/d2o_sld_spread.png}}\\%
\caption{a) Reflectivity from a DMPC bilayer measured at three contrasts, with median in red, and 500 samples from the posterior distribution. Dataset for \ce{Si} match offset by 0.1, \ce{H_2O} contrast offset by 0.001. b) SLD profile of the \ce{D_2O} model showing 500 samples from the posterior distribution, as well as the median.}
\end{figure}

@@

\section*{Conclusions}\label{conclusions}
\refnx\ is a powerful tool for least-squares or Bayesian analysis of neutron and X-ray reflectometry data that is ideally usable for reproducible research with \Jupyter\ notebooks, and with extensibility in mind. Its features include: MCMC sampling of posterior distribution for parameters, structural models constructed from modular components with physically relevant parameterisation, algebraic inter-parameter constraints, mixed area models, co-refinement of multiple datasets, probability distributions for parameter bounds used directly for log-prior terms, and a (\Jupyter\) \ipywidgets\ GUI.

\section*{Acknowledgements}
We acknowledge Anton Le Brun (ANSTO) for the provision of the lipid bilayer datasets in the example, James Hester (ANSTO) for comments made on the draft manuscript, and Andrew McCluskey (Bath University) and Isaac Gresham (UNSW) for important feedback on refnx development.
